Local PCA Shows How the Effect of Population Structure Differs Along the Genome

### 2018-08-21 writeup/local_pca_paper.tex
run_016913 simulation added
discussion of tests

\paragraph{Testing}
Figure \ref{fig:test_missing} shows tests with lots of missing data.
Also the covariance matrix was chosen to have the top two eigenvalues equal,
so that eigenvector switching is expected.

\subsection{Visualization of results}

@@

@@

Address
mutation rate variation;
recombination rate variation;
choice of number of PCs;
choice of window size;
variation in missingness;
maybe differing sample sizes and reweighting.

@@

\begin{figure}
    \begin{center}
       \includegraphics[width=0.82\textwidth]{FigS_plot_corners-1_run_012720}
    \end{center}
    \caption{
        MDS visualizations of simulation run 012720 (RECOMBTYPE, ``step"; SELTYPE, ``neutral")
        \label{sfig:mds_run_012720}
    }
\end{figure}

\begin{figure}
    \begin{center}
       \includegraphics[width=0.82\textwidth]{FigS_plot_corners-1_run_027034}
    \end{center}
    \caption{
        MDS visualizations of simulation run 027034 (RECOMBTYPE, ``base"; SELTYPE, ``neutral")
        \label{sfig:mds_run_027034}
    }
\end{figure}

\begin{figure}
    \begin{center}
       \includegraphics[width=0.82\textwidth]{FigS_plot_corners-1_run_015598}
    \end{center}
    \caption{
        MDS visualizations of simulation run 015598 (RECOMBTYPE, ``hotspot"; SELTYPE, ``neutral")
        \label{sfig:mds_run_015598}
    }
\end{figure}

\begin{figure}
    \begin{center}
       \includegraphics[width=0.82\textwidth]{FigS_plot_corner_pca-1_run_015598}
    \end{center}
    \caption{
        PCA plots for regions colored in \Figure \ref{sfig:mds_run_015598}
        \label{sfig:pca_corner_run_015598}
    }
\end{figure}

@@

\begin{figure}
    \begin{center}
       \includegraphics[width=0.82\textwidth]{FigS_plot_corners-1_run_031486}
    \end{center}
    \caption{
        MDS visualizations of simulation run 031486 (RECOMBTYPE, ``base"; SELTYPE, ``selected")
        \label{sfig:mds_run_031486}
    }
\end{figure}

\begin{figure}
    \begin{center}
       \includegraphics[width=0.82\textwidth]{FigS_plot_corners-1_run_009673}
    \end{center}
    \caption{
        MDS visualizations of simulation run 009673 (RECOMBTYPE, ``step"; SELTYPE, ``selected")
        \label{sfig:mds_run_009673}
    }
\end{figure}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Visualization of results}

@@

%%%%%%%%%%%%%%%%%%%
\subsection{Testing}

We tested the method using two types of simulation.
First, to verify expected behavior, we simulated ``genomes'' as an independent sequence
of correlated Gaussian ``genotypes'', using a different covariance matrix
in the first quarter, middle half, and last quarter of the chromosome.
To verify robustness to missing data,
we ran the method after randomly dropping 50\% of the genotypes in the first half of the genome;
if the method is misled by missing data, then it will distinguish the two halves of the chromosome
rather than the segments having different covariance matrices.
Details are given in Appendix \ref{apx:gaussian_sims}.

To provide a realistic test,
we next used forwards-time, individual-based simulations, implemented using SLiM v3 \citep{slim_v3},
which are described in detail in Appendix \ref{apx:slim_sims}.
To provide realistic population structure for PCA to identify,
each simulation ahd at least 5,000 diploid individuals,
living across a continuous square range,
with Gaussian dispersal and local density-dependent competition.
Each genome was modeled on human chromosome 7, which is $1.54 \times 10^8$bp long,
with an overall recombination rate of 1.6785 crossovers per chromosome per generation.
To improve speed, we used tskit \citep{tskit_paper} to record tree sequences
and add neutral mutations afterwards, at a rate of $10^{-9}$ per bp per generation.
Most simulations were neutral, but we also included linked selection, of two types.
First, % run_031486
we introduced selected mutations into two regions, 
which extended from 1/3 to 1/2 and from 5/6 to the end of the genome respectively.
These had selection coefficients from a Gamma distribution with shape 2 and mean 0.005
at a rate of $10^{-10}$ per bp, that were either beneficial (with probability $1/30$)
or deleterious (otherwise).
Second, % run_016913
to roughly model a recent expansion followed by local adaptation,
we introduced mutations in the same manner as above,
except that mutations were no longer unconditionally deleterious or beneficial:
each selection coefficient was multiplied by a factor depending on the spatial location
of the individual being evaluated, varying linearly from -1 at the left side of the range
to +1 at the right edge.
In all simulations, genome-wide PCA displayed a map of the population range, as expected.

@@

\plr{Address:
    mutation rate variation;
    recombination rate variation;
    choice of number of PCs;
    choice of window size;
    variation in missingness;
    maybe differing sample sizes and reweighting.  }

Simple non-population-based simulations with Gaussian ``genotypes''
showed that the method performs as expected, clearly separating regions of the genome
with different underlying covariance matrices
without being affected by extreme differences in amount of missing data
(Supplemental Figure \ref{fig:test_missing}).
This simulation also verifies insensitivity to ordering of top PCs,
since it was performed using a covariance matrix with the top two eigenvalues equal,
so that the order of empirical eigenvectors (PCs) switches randomly.
The same type of simulations with dramatically different sample sizes
showed that using weighted PCA substantially improved identification of different genomic regions
(Supplemental Figure \ref{fig:test_weighted_pca}).

Individual-based simulations using SLiM \citep{slim_v3} allowed us to test the effects
of recombination and mutation rate variation, as well as linked selection.
Results are shown in Supplemental Figures \ref{sfig:slim_sims1} and \ref{sfig:slim_sims2}.
As expected, varying recombination rate stepwise by a factor of 64
did not induce patterns in the MDS visualizations correlated with recombination rate.
Since varying mutation rate with a fixed recombination map
is equivalent to varying the recombination map and remapping windows,
we used the same simulations to show that variation in mutation rate similarly has no effect.
On the other hand,
a recombination map with hotspots
(the HapMap human female map for chromosome 7 \citep{hapmap})
induced outliers at long regions of low recombination rate
(also as expected).

Simulations with linked selection produced mixed results
(Figure \ref{sfig:linked_selection}).
The method strongly identified the regions under spatially varying linked selection.
It also identified the regions (although less unambiguously)
with constant selection and stepwise varying recombination rate,
but did not clearly identify them with constant recombination rate.
This difference is likely because recombination rates are overall lower in the first case,
leading to a stronger effect of linked seleciton.
These tests are not meant to be comprehensive survey of linked selection,
but only to demonstrate that linked selection can produce signals simlar to what we see in real data.

@@

\begin{figure}
    \begin{center}
       \includegraphics[width=0.82\textwidth]{FigS_plot_corners-1_run_027034}
       \includegraphics[width=0.82\textwidth]{FigS_plot_corners-1_run_012720}
       \includegraphics[width=0.82\textwidth]{FigS_plot_corners-1_run_015598}
       \includegraphics[width=0.82\textwidth]{FigS_plot_corner_pca-1_run_015598}
    \end{center}
    \caption{
        MDS visualizations of the results of individual-based simulations using SLiM
        (see Appendix \ref{apx:slim_sims} for details).
        All simulations are neutral, and recombination is:
        \textbf{(top)} constant;
        \textbf{(top middle)} varies stepwise by factors of two in seven equal-length segments,
        with highest rates on the ends, so the middle segment has a recombination rate 64 times lower
        than the ends;
        \textbf{(bottom middle)} according to the HapMap human female chromosome 7 map.
        The \textbf{bottom} figure shows PCA maps corresponding to the three colored windows
        of the last (HapMap) situation;
        the outlying regions are long regions of low recombination rate,
        so that region can be dominated by a few correlated trees,
        similar to an inversion.
    } \label{sfig:slim_sims1}
\end{figure}

\begin{figure}
    \begin{center}
    \end{center}
    \caption{
        PCA plots for regions colored in \Figure \ref{sfig:mds_run_015598}
        \label{sfig:pca_corner_run_015598}
    }
\end{figure}

@@

\begin{figure}
    \begin{center}
       \includegraphics[width=0.82\textwidth]{FigS_plot_corners-1_run_031486}
    \end{center}
    \caption{
        \plr{KEEP}
        MDS visualizations of simulation run 031486 (RECOMBTYPE, ``base"; SELTYPE, ``selected")
        \label{sfig:mds_run_031486}
    }
\end{figure}

\begin{figure}
    \begin{center}
       \includegraphics[width=0.82\textwidth]{FigS_plot_corners-1_run_009673}
    \end{center}
    \caption{
        \plr{KEEP}
        MDS visualizations of simulation run 009673 (RECOMBTYPE, ``step"; SELTYPE, ``selected")
        \label{sfig:mds_run_009673}
    }
\end{figure}

\begin{figure}
    \begin{center}
       \includegraphics[width=0.82\textwidth]{FigS_plot_corners-1_run_016913}
    \end{center}
    \caption{
        MDS visualizations of simulation run 009673 (RECOMBTYPE, ``base"; SELTYPE, ``balancing")
        \label{sfig:mds_run_016913}
    }
\end{figure}

\begin{figure}
    \begin{center}
       \includegraphics[width=0.82\textwidth]{FigS_plot_corner_pca-1_run_016913}
    \end{center}
    \caption{
        PCA plots for regions colored in \Figure \ref{sfig:mds_run_016913}
        \label{sfig:pca_corner_run_016913}
    }
\end{figure}
